Is essentiality of genes conserved in Enterobacteriaceae?

### 2016-10-11 documents/manuscript.tex
Added the difference between different core-essential definition methods, added the difference core-essential genes in root and branches

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Is the essentiality of genes conserved in Enterobacteriaceae?}
\newline{Is gene essentiality conserved?}
\newline{Are essential genes conserved?}
}
\newline
% authors go here:
\\
%Fatemeh Ashari Ghomi\textsuperscript{1},
%Paul P. Gardner\textsuperscript{1,*},
%Lars Barquist\textsuperscript{2}
%\\
%\bigskip
%\bf{1} University of Canterbury
%\\
%\bf{2} Wurzburg University
%\\
%\bigskip
%* paul.gardner@canterbury.ac.nz

@@

\section{Transposon mutagenesis}

{\bf overview paragraph.}

We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}. These strains are all selected from Enterobacteriaceae family. Enterobacteriaceae is a family that includes Gram-negative bacteria with different host ranges and pathogenecity found in soil, water, plants, animals and humans \cite{brenner_bergeys_2006}. In humans, various strains from this family can cause diarrhoea, septicaemia, urinary tract infection, meningitis, respiratory disease, and wound and burn infection \cite{brenner_bergeys_2006}. Besides, they can infect poultry and livestocks and cause financial losses for farmers \cite{brenner_bergeys_2006}. Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.

We have used Tn5 transposon to generate single inserted mutants and placed our mutants in a selective media for Tn5. We have picked the mutants and pooled them and used the splinkerette adapter and primers designed in \cite{barquist_tradis_2016} for PCR enrichment. Then we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome. The number of insertions in a gene can imply the degree of essentiality for that gene.

@@

We have divided the genes into three groups: essential genes, ambiguous, non-essential genes, and beneficial losses. We have adapted the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes. The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}. The first peak shows the genes with no or just a few insertions which are considered as essential genes. We have fitted an exponential distribution to the first peak and a gamma distribution to the second one. Then, we have calculated the log odds ratio for belonging to each of these distributions for each gene. The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential. Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.

@@

%\section{Are transposons biased towards certain positions?}
\section{Are there biases in transposon mutagenesis data?}

{\bf OVERVIEW PARAGRAPH:}

Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}. We have performed a detailed study of these biases. {\bf LIST TYPES OF BIAS HERE.}

{\bf ORIGIN OF REPLICATION BIAS:} To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome. Fig.\@ \ref{fig:distance-bias} shows the results. The red line is a loess curve that has been fitted to the data when the smoothness parameter equals $0.2$. The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication. A possible explanation for this phenomenon is that the bacteria were under replication while being infected with transposons. Therefore, the number of gene copies close to the origin of replication was greater and more insertions have occurred in these genes. To overcome this bias, we have normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplying this value by the average insertion index.

@@

>>>>

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{Are essential genes conserved?}
}
\newline
% authors go here:
\\
%Fatemeh Ashari Ghomi\textsuperscript{1},
%Paul P. Gardner\textsuperscript{1,*},
%Lars Barquist\textsuperscript{2}
%\\
%\bigskip
%\bf{1} University of Canterbury
%\\
%\bf{2} Wurzburg University
%\\
%\bigskip
%* paul.gardner@canterbury.ac.nz

@@

\section{Transposon mutagenesis}
During evolution species can gain or lose genes. We have investigated if these gene gain and losses are related to the essentiality of the genes themselves. For this, we have evaluated the essentiality of genes and their conservation and the relationship between these two. We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}. These strains are all selected from Enterobacteriaceae family. Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.

Transposon mutagenesis is frequently used for the study of essentiality. We have used Tn5 transposon to generate single inserted mutants and placed our mutants in a selective media for Tn5. We have picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}. Then, we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome. The number of insertions in a gene implies the degree of essentiality for that gene.

@@

We have divided the genes into three groups: essential genes, ambiguous, non-essential genes, and beneficial losses. We have utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes. The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}. The first peak shows the genes with no or just a few insertions which are considered as essential genes. We have fitted an exponential distribution to the first peak and a gamma distribution to the second one. Then, we have calculated the log odds ratio for belonging to each of these distributions for each gene. The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential. Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.

@@

%\section{Are transposons biased towards certain positions?}
\section{Are there biases in transposon mutagenesis data?}
To evaluate the essentiality of a gene, we use the number of insertions within that gene. However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis. Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}. We have performed a detailed study of these biases. The biases that we have studied include: gene length bias, origin of replication bias, preferred insertion motif bias, and positional bias within genes.

\subsection{Length bias}
Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions. As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes. 
In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index. It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
\begin{figure*}
\centering
\includegraphics[page=42, scale=0.3]{biases.pdf}
\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
\label{fig:length-bias}
\end{figure*}

\subsection{Origin of replication bias}
To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome. Fig.\@ \ref{fig:distance-bias} shows the results. The red line is a loess curve that has been fitted to the data when the smoothness parameter equals $0.2$. The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication. A possible explanation for this phenomenon is that the bacteria were under replication while being infected with transposons. Therefore, the number of gene copies close to the origin of replication was greater and more insertions have occurred in these genes. To overcome this bias, we have normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplying this value by the average insertion index.

@@

As these three methods lead to different results, we have compared the differences between the genes found in these three methods. For this, we have compared the set of core essential genes resulted from intersection and ancestral insertion index methods and performed word enrichment analysis that was explained before on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method. Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure. The results are depicted in Fig.\@\ref{fig:method-difference}.

\begin{figure*}
\centering
\begin{tabular}{c c}
\includegraphics[scale=0.4]{ancestral-intersection-pval.pdf}&
\includegraphics[scale=0.4]{dollo-ancestral-pval.pdf}\\
\includegraphics[scale=0.4]{dollo-intersection-pval.pdf}&
\end{tabular}
\caption{The figure shows the difference between core essential genes in intersection and ancestral insertion index methods, ancestral insertion index and fdollop methods, and intersection and fdollop methods. The red line shows P-value = 0.05. The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure. The top left figure shows words that are enriched in core essential genes found using ancestral insertion index method but not enriched in core essential genes found using intersection method. The top right figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using ancestral insertion index method. The bottom left figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using intersection method.}
\label{fig:method-difference}
\end{figure*}
